---
title: "MAGMA - MAny Group MAtching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MAGMA - MAny Group MAtching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

```{r setup, include = FALSE}
library(MAGMA)
```

## Introduction

The MAGMA-package offers nearest neighbor matching for two to four groups. It also includes the option to match data of a 2x2 Design. In addition, MAGMA contains a framework to evaluate post matching balance. This vignette is a tutorial of MAGMA. It shows the main MAGMA function on an included simulated dataset. Overall, this vignette includes a two-group, a three-group, and a 2x2-design which is also approximately equivalent to a four-group example. This three examples are conducted in a standard matching process as well as with exact matching. Parts of this tutorial are also included on the respective help pages of the functions. For details of the algorithm and a more extensive introduction of MAGMA, see Feuchter et al. (2022).

## The MAGMA Simulated Dataset

This simulated dataset contains 800 cases and 14 variables. It is available after loading the MAGMA package. Associations between the variables were modeled to increase usability and comprehensibility of this tutorial. Any associations are, therefore, not interpretable. For information regarding this simulated dataset use `?MAGMA_sim_data`. Below, you find a short glimpse on the data and its structure

```{r data_introduction}
str(MAGMA_sim_data)

head(MAGMA_sim_data)
```
Note that the three last variables are propensity scores estimated with the `twang` package (Ridgeway et al., 2015). They serve as distance indicators for the three examples. We recommend to use propensity scores as distance input for MAGMA.
Before we start with the first example, we investigate the descriptive statistics of the simulated dataset.

```{r data_descriptives}
MAGMA_sim_data %>%
  select(sex, gifted_support, teacher_ability_rating, enrichment,
         parents_academic, GPA_school, IQ_score, Motivation, college_GPA) %>%
  psych::describe()

MAGMA_sim_data %>%
  select(sex, gifted_support, teacher_ability_rating, enrichment,
         parents_academic, GPA_school, IQ_score, Motivation, college_GPA) %>%
  cor()
```


## Two-Group Example

This is a fictional two-group example. The independent variable (i.e., the treatment) is `gifted_support`. We are interested in how receiving of giftedness support affects the college GPA. However, other variables, such as intelligence are associated with both, giftedness support and the college GPA. This is because gifted students were chosen to participate in this support program based on their previous achivement or intelligence test results. Therefore, we aim to reduce the effect of these covariates through the matching. For this example, the covariates are the *high school GPA*, *intelligence*, *motivation*, *sex*, and whether at least one parent has an *academic background*. Note that MAGMA can only consider metric and binary variables for balance estimation.

```{r covariates_gifted}
covariates_gifted <- c("GPA_school", "IQ_score", "Motivation", "parents_academic", "sex")
```

After this definition we can estimate the initial unbalance over all cases in our dataset. Here, the function `initial_unbalnce` estimates the four balance criteria with the whole sample. More specific, it indicates the unbalance in the sample before matching. Therefore, it is necessary to define the dataset (`MAGMA_sim_data`), the grouping variable (`gifted_support`), and all binary and metric covariates of interest (`covariates_gifted`). This results in a vector with the length four - the four balance criteria.

```{r unbalance_gifted}
unbalance_gifted <- initial_unbalance(data = MAGMA_sim_data,
                                      group = "gifted_support",
                                      covariates = covariates_gifted)
unbalance_gifted
```
As may be expected from the descriptive statistics, the cases that received giftedness support differed at least on some variables substantially from the cases that did not receive giftedness support. In the next step, we compute the matching. This process is based on the respective propensity score, namely `ps_gifted`. We first address standard matching and deal after that with exact matching.

### Standard Matching

For standard nearest neighbor matching we use the function `MAGMA`. This function has four arguments. The first specifies the dataset (`MAGMA_sim_data`), the second the name of the grouping variable as a character (`"gifted_support"`), and the third the name of the distance variable/propensity score as a character (`"ps_gifted"`). The fourth argument is optional. Since MAGMA matching has some computational load it includes the option of parallel computation. Per default, only one core is used. If parallel computation is desired, a larger than one integer can be defined for the `cores` argument. If the specified value exceeds the number of available cores, MAGMA sets this argument to the maximum available cores.

Applying this function results in a dataset, that augments the original dataset for three variables. The first, `weight`, indicates whether this case was matched or not. The second, `step`, names the iterative step in which a case was matched. More specifically, the cases with the smallest distance are matched first and, thus, receive `1` in the step variable. The matched cases with the second smallest distance receive the `2` and so on. The third variable, `distance` specifies this distance of the matched cases.

```{r standard_2_group_matching}
MAGMA_sim_data_gifted <- MAGMA(Data = MAGMA_sim_data,
                                group = "gifted_support",
                                dist = "ps_gifted",
                                cores = 2)
str(MAGMA_sim_data_gifted)
```

#### Balance estimation

MAGMA matches cases iteratively until all cases of the smallest group are matched. Thus, it is necessary to find the optimal sample size after matching. To do so, `Balance_MAGMA` estimates the four balance criteria for each possible sample size. Note that MAGMA uses a lower limit of at least *n* = 20 per group. To apply the function you need at least to specify your dataset (`MAGMA_sim_data_gifted`), the name of the grouping variable as a character (`"gifted_support"`), and your covariates as a character vector (`covariates_gifted`). Moreover, `Balance_MAGMA` needs the name of the variable, that indicates the iterative `step` in that a case was matched. Since the `MAGMA` function names this variable `"step"` this name is the default of the `step` argument. If you want to use `Balance_MAGMA` on a dataset with various MAGMA-matchings or to evaluate balance of another matching algorithm, you need to define the respective name as a character.

`Balance_MAGMA` returns a list length four. In the case of a univariate grouping variable (compare 2x2 matching), this list contains three numeric vectors and one list length two. The three vectors form Pillai's Trace, mean *g*, and adjusted *d*-ratio over the iteratively increased sample size. The list length two includes one numeric vector, *d*-ratio, and one matrix. The matrix specifies all pairwise *d*'s for all covariates. Both, *d*-ratio and pairwise *d*'s, are also estimated teratively over the increased sample size. The example below estimates the Balance for the matching and shows the balance criteria and pairwise effects for a group sample size of *n* = 100.

```{r Balance_standard_2_group_matching}
Balance_gifted <- Balance_MAGMA(data = MAGMA_sim_data_gifted,
                                group = "gifted_support",
                                covariates = covariates_gifted,
                                step = "step") #Not necessary to define here

str(Balance_gifted) #NA's result of minimum required sample size

#Balance criteria for a 100 cases per group
Balance_100_gifted <- c(Balance_gifted$Pillai[100],
                        Balance_gifted$d_ratio$d_rate[100],
                        Balance_gifted$mean_effect[100],
                        Balance_gifted$adjusted_d_ratio[100])
Balance_100_gifted

Balance_gifted$d_ratio$effects[, 100]
```

#### Balance plotting

With this information regarding the balance with respect to sample size, the desired sample size can be selected. The "optimal" sample size does strongly depend on the specific research aim. As example, sample size requirements of some statistical procedures may narrow the selection, or the need to eliminate a specific pairwise effect. We, therefore, want to encourage to adapt the selection of optimal sample size to the specific research aim.

A first indicator on such an optimal model is the trend of the balance criteria across the iteratively increased sample size. The function `Plot_MAGMA` visualize this trend. The one mandatory argument, `Balance`, is specifying the object of the balance estimation results. Note that this object needs to be the result of a `Balance_MAGMA` function. Per default, `Plot_MAGMA` creates plots for all four balance criteria. However, the `criterion` argument enables the plotting of specific balance criteria. Note that running this code leads to a warning. This warning is a result of the specified minimum sample size for balance estimation and can be ignored.

The trend of this plots strongly depends on the data. Nonetheless, they can be a good indicator, where breakpoints in the sample size-balance associations may be located. `Plot_MAGMA` uses fixed limits for the y-axis.This ensures comparability of this plots and prevents misinterpretation due to a restricted y-axis range. Both, *d*-ratio and adjusted *d*-ratio, vary between 0 and 1, wherefore those form the limits for these criteria. Although Pillai's Trace has theoretically the same range, we set the limits to 0 and 0.5. Larger values are unlikely and indicate a bad balance. A stricter limitation could restrict a comprehensive trend evaluation. For mean *g* we set the limits to 0 and 1. The lower limit 0 corresponds to the theoretical lower limit. The rationality for the upper limit is comparable to the one of Pillai's Trace.

```{r Plots_standard_2_group_matching, fig.height = 5, fig.width = 7.5}
Plot_MAGMA(Balance = Balance_gifted,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted
```

#### Balance table

Besides the trend over sample size the absolutely best values for each criterion are of interest. The function `Table_MAGMA` extracts these values and returns them in an APA table in an extra file. More specific it returns a 4x5 table. In each of the four rows one balance criterion has its optimal value. The five columns represent the sample size per group, where the respective criterion is optimal, and the four balance criteria with this sample size per group. Besides creating a file with an APA table, the table is printed to the console, too.

Similar to `Plot_MAGMA` the first argument `Balance` of `Table-MAGMA` specifies the object of the balance estimation results. The second argument `filename` indicates the desired name of the extra file with the APA table. This argument is a character.

```{r Table_standard_2_group_matching}
Table_MAGMA(Balance = Balance_gifted,
            filename = "Balance_gifted.docx")
```

### Exact Matching

In comparison to standard matching, only cases sharing the same value on a defined variable can be matched in exact matching. In this example, the variable for exact matching is `enrichment`. This means, that cases that participated in afternoon enrichment can only be matched with other cases that participated, and vice versa. Reasons to select exact matching instead of standard matching may be, for example, a nested data structure. Besides this preselection of possible matches the process of matching stays the same. The main differene is that we need to us `MAGMA_exact` instead of `MAGMA` to match the data. This function has a further argument, where the name of the exact variable needs to be defined as a character (`"enrichment"`). Note that the initial unbalance is the same as for standard matching.

```{r exact_2_group_matching}
MAGMA_sim_data_gifted_exact <- MAGMA_exact(Data = MAGMA_sim_data,
                                           group = "gifted_support",
                                           dist = "ps_gifted",
                                           exact = "enrichment",
                                           cores = 2)
str(MAGMA_sim_data_gifted_exact)
```

#### Balance Estimation and Visualization

The steps of balance estimation and visualization are the same as for standard matching. Therefore, we summarize these steps. The only change we need to consider is the different data frame that contains the result of the exact matching

```{r Balance_exact_2_group_matching, fig.height = 5, fig.width = 7.5}
Balance_gifted_exact <- Balance_MAGMA(data = MAGMA_sim_data_gifted_exact,
                                      group = "gifted_support",
                                      covariates = covariates_gifted,
                                      step = "step") #Not necessary to define here

str(Balance_gifted_exact)

#Balance criteria for a 100 cases per group
Balance_100_gifted_exact <- c(Balance_gifted_exact$Pillai[100],
                              Balance_gifted_exact$d_ratio$d_rate[100],
                              Balance_gifted_exact$mean_effect[100],
                              Balance_gifted_exact$adjusted_d_ratio[100])
Balance_100_gifted_exact

Balance_gifted_exact$d_ratio$effects[, 100]

#Ploting trend over increasing samle size
Plot_MAGMA(Balance = Balance_gifted_exact,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted

#Creating APA table
Table_MAGMA(Balance = Balance_gifted_exact,
            filename = "Balance_gifted_exact.docx")

```

## Three-Group Example

This is a fictional three-group example. The independent variable (i.e., the treatment) is `teacher_ability_rating`. We are interested in how teacher rated ability, which ranged from below average to average and above average, affects the college GPA. While the process of matching and balance evaluation is the same as in the two-group example, the grouping variable (`teacher_ability_rating`) and consequently the distance variable/propensity score changes (`ps_tar`). Moreover, we slightly change the covariates by substituting `sex` through `gifted_support`.

```{r covariates_tar}
covariates_tar <- c("GPA_school", "IQ_score", "Motivation", "parents_academic", "gifted_support")
```

### Standard Matching

Due to this changes in grouping variable and covariates, we need to estimate the initial unbalance for teacher rated ability. Note that due to having three groups the estimation of mean *g* slightly changes. However, this does only affect the backend computation, while the code and the display of results are not affected. For more details, see Feuchter et al. (2022).

```{r unbalance_tar}
unbalance_tar <- initial_unbalance(data = MAGMA_sim_data,
                                   group = "teacher_ability_rating",
                                   covariates = covariates_tar)
unbalance_tar
```

After this check of initial unbalance we can conduct the matching. As mentioned, we need to adapt the grouping variable and the distance variable/propensity score to this example.

```{r standard_3_group_matching}
MAGMA_sim_data_tar <- MAGMA(Data = MAGMA_sim_data,
                            group = "teacher_ability_rating",
                            dist = "ps_tar",
                            cores = 2)
str(MAGMA_sim_data_tar)
```

#### Balance Estimation and Visualization

We only need to change the specific object names, covariates, grouping variable, and the filename for the APA table. Besides this, the process of balance estimation and visualization remains the same.

```{r Balance_3_group_matching, fig.height = 5, fig.width = 7.5}
Balance_tar <- Balance_MAGMA(data = MAGMA_sim_data_tar,
                                      group = "teacher_ability_rating",
                                      covariates = covariates_tar,
                                      step = "step") #Not necessary to define here

str(Balance_tar)

#Balance criteria for a 100 cases per group
Balance_100_tar <- c(Balance_tar$Pillai[100],
                     Balance_tar$d_ratio$d_rate[100],
                     Balance_tar$mean_effect[100],
                     Balance_tar$adjusted_d_ratio[100])
Balance_100_tar

Balance_tar$d_ratio$effects[, 100]

#Ploting trend over increasing samle size
Plot_MAGMA(Balance = Balance_tar,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted

#Creating APA table
Table_MAGMA(Balance = Balance_tar,
            filename = "Balance_tar.docx")

```

### Exact Matching

We use the same covariates as we used for the standard three-group matching. The only change constitutes using `sex` as exact variable. Attached you find the whole matching, balance estimataion, and visualization process.

```{r exact_3_group_matching, fig.height = 5, fig.width = 7.5}
MAGMA_sim_data_tar_exact <- MAGMA_exact(Data = MAGMA_sim_data,
                                        group = "teacher_ability_rating",
                                        dist = "ps_tar",
                                        exact = "sex",
                                        cores = 2)
str(MAGMA_sim_data_tar_exact)

Balance_tar_exact <- Balance_MAGMA(data = MAGMA_sim_data_tar_exact,
                                   group = "teacher_ability_rating",
                                   covariates = covariates_tar,
                                   step = "step") #Not necessary to define here

str(Balance_tar_exact)

#Balance criteria for a 100 cases per group
Balance_100_tar_exact <- c(Balance_tar_exact$Pillai[100],
                           Balance_tar_exact$d_ratio$d_rate[100],
                           Balance_tar_exact$mean_effect[100],
                           Balance_tar_exact$adjusted_d_ratio[100])
Balance_100_tar_exact

Balance_tar_exact$d_ratio$effects[, 100]

#Ploting trend over increasing samle size
Plot_MAGMA(Balance = Balance_tar_exact,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted

#Creating APA table
Table_MAGMA(Balance = Balance_tar_exact,
            filename = "Balance_tar_exact.docx")
```

## 2x2/four-group Example

In this fictional 2x2-design example the variables `gifted_support` and `enrichment` are the independent (i.e., the treatment) variables. We are interested in how the receiving of giftedness support and the participation in afternoon enrichment affects the college GPA. The process of matching and balance evaluation is the same as in the former examples. The grouping variable consists of two variables, wherefore we use a character vector as input for the groupung variable (`c("gifted_support", "enrichment")`). Consequently, we also adapt the distance variable/propensity score (`ps_2x2`). Moreover, we use the same covariates as in the two-group example. Note that this 2x2 matching is nearly equivalent to a four group matching using a multinominal variable indicating the four cells of the 2x2 group design. The multinominal variable for this example is `support_enrichment`. Major distinctions between such a multinominal four-group matching and a 2x2 matching occur only for balance estimation. This differences are described more thoroughly below.

```{r covariates_2x2}
covariates_2x2 <- c("GPA_school", "IQ_score", "Motivation", "parents_academic", "sex")
```

### Standard Matching

Due to this changes in grouping variable and covariates, we need to estimate the initial unbalance for the two independent variables `gifted_support` and `enrichment`. The result of the `initial_unbalance` function slightly changes in this two-factorial design. Pillai's Trace is estimated separately for the two main effects and the interaction. As can be seen, *d*-ratio, mean *g*, and adjusted *d*-ratio are the same for a 2x2 or a four-group matching. Changes only occur for Pillai's Trace. 

```{r unbalance_2x2}
unbalance_2x2 <- initial_unbalance(data = MAGMA_sim_data,
                                   group = c("gifted_support", "enrichment"),
                                   covariates = covariates_2x2)
unbalance_2x2

unbalance_4_group <- initial_unbalance(data = MAGMA_sim_data,
                                   group = "support_enrichment",
                                   covariates = covariates_2x2)
unbalance_4_group
```

After this check of initial unbalance we can conduct the matching. As mentioned, we need to adapt the grouping variable and the distance variable/propensity score to this example. Moreover, the conducting of a 2x2/four-group matching has some computational load. If a RAM threshold is exceeded, MAGMA computes quasi-systematic matching. The rationality of this is described elsewhere. Again, this only affects the computational backend and does not affect the user application. Note that in the case of a multifactorial matching, MAGMA returns a fourth variable. This fourth variable is a multinominal version of the two independent variables (`group_long`).

```{r standard_2x2_group_matching}
MAGMA_sim_data_2x2 <- MAGMA(Data = MAGMA_sim_data,
                            group = c("gifted_support", "enrichment"),
                            dist = "ps_2x2",
                            cores = 2)
str(MAGMA_sim_data_2x2)

MAGMA_sim_data_four_group <- MAGMA(Data = MAGMA_sim_data,
                            group = "support_enrichment",
                            dist = "ps_2x2",
                            cores = 2)
str(MAGMA_sim_data_four_group)

#Making data frames comparable
MAGMA_sim_data_2x2_comp <- MAGMA_sim_data_2x2 %>%
  select(step, weight, distance)

MAGMA_sim_data_four_group_comp <- MAGMA_sim_data_four_group %>%
  select(step, weight, distance)

#ColSums equal the number of cases
identical(MAGMA_sim_data_2x2_comp, MAGMA_sim_data_four_group_comp)
```

#### Balance Estimation and Visualization

For this example, we only continue with the 2x2 design for balance estimation. The only difference to the former examples in the process is again, that the grouping variable is a character vector length two. As a result of this, `Plot_MAGMA` displays three Pillai's Trace plots corresponding to the two main effects and the interactin respectively.

```{r Balance_2x2_group_matching, fig.height = 5, fig.width = 7.5}
Balance_2x2 <- Balance_MAGMA(data = MAGMA_sim_data_2x2,
                             group = c("gifted_support", "enrichment"),
                             covariates = covariates_2x2,
                             step = "step") #Not necessary to define here

str(Balance_2x2)

#Balance criteria for a 100 cases per group
Balance_100_2x2 <- c(Balance_2x2$Pillai[1, 100],
                     Balance_2x2$Pillai[2, 100],
                     Balance_2x2$Pillai[3, 100],
                     Balance_2x2$d_ratio$d_rate[100],
                     Balance_2x2$mean_effect[100],
                     Balance_2x2$adjusted_d_ratio[100])
Balance_100_2x2

Balance_2x2$d_ratio$effects[, 100]

#Ploting trend over increasing samle size
Plot_MAGMA(Balance = Balance_2x2,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted

#Creating APA table
Table_MAGMA(Balance = Balance_2x2,
            filename = "Balance_2x2.docx")

```

### Exact Matching

We use the same covariates as we used for the standard 2x2 matching. The only change constitutes using `teacher_ability_rating` as exact variable. Attached you find the whole matching, balance estimataion, and visualization process.

```{r exact_2x2_group_matching, fig.height = 5, fig.width = 7.5}
MAGMA_sim_data_2x2_exact <- MAGMA_exact(Data = MAGMA_sim_data,
                                        group = c("gifted_support", "enrichment"),
                                        dist = "ps_2x2",
                                        exact = "teacher_ability_rating",
                                        cores = 2)
str(MAGMA_sim_data_2x2_exact)

Balance_2x2_exact <- Balance_MAGMA(data = MAGMA_sim_data_2x2_exact,
                                   group = c("gifted_support", "enrichment"),
                                   covariates = covariates_2x2,
                                   step = "step") #Not necessary to define here

str(Balance_2x2_exact)

#Balance criteria for a 100 cases per group
Balance_100_2x2_exact <- c(Balance_2x2_exact$Pillai[1, 100],
                           Balance_2x2_exact$Pillai[2, 100],
                           Balance_2x2_exact$Pillai[3, 100],
                           Balance_2x2_exact$d_ratio$d_rate[100],
                           Balance_2x2_exact$mean_effect[100],
                           Balance_2x2_exact$adjusted_d_ratio[100])
Balance_100_2x2_exact

Balance_2x2_exact$d_ratio$effects[, 100]

#Ploting trend over increasing sample size
Plot_MAGMA(Balance = Balance_2x2_exact,
           criterion = c("Pillai", "d_ratio", "mean_g", "Adj_d_ratio")) #Could be omitted

#Creating APA table
Table_MAGMA(Balance = Balance_2x2_exact,
            filename = "Balance_2x2_exact.docx")
```

## References

Feuchter, M. D., Urban, J., Scherrer, V., Breit, M., & Preckel, F. (2022). Introduction and Demonstration of the Many-Group Matching  (MAGMA)-Algorithm: Matching Solutions for Two or More Groups. https://doi.org/10.17605/OSF.IO/AEDXB.

Ridgeway, G., McCaﬀrey, D., Morral, A., Burgette, L., & Griﬃn, B. A. (2015). The twang 
package (1.6). https://CRAN.R-project.org/package=twang 
